---
title: "Analyzing UK Price Data visualizations"
author: "Andres Felipe Virgüez"


output: html_document
---

**Seet up of the project**

```{r, warning=FALSE, message=F}
# Clear the enviroment
rm(list = ls())

# Load or install libraries
libraries_list = c("tidyverse", "leaflet", "dygraphs", "xts", "sp", "sf", "plotly", "rgdal", "geojsonio", "viridis")
new_librariss <- libraries_list[!(libraries_list %in% installed.packages()[,"Package"])]
if(length(new_librariss)) install.packages(new_librariss) # Aqui instalo las librerias que no estan
lapply(libraries_list, library, character.only = TRUE) # Aqui cargo los paquetes


# Set the current directory as the working directory 
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

Loading the data...

```{r, warning=FALSE, message=F}

df = read_csv("data/ppdata_lite.csv")

```


### What are the most and less expensive London Boroughs? 

```{r, warning=FALSE, message=F}
london_df = df %>% filter(county == 'GREATER LONDON') # Data set for only London
london_df_2015 = london_df %>% filter(format(date_of_transfer, format = "%Y") == 2015) # Data set for only London
print(length(unique(london_df$district))) # Print the number of districts also called Boroughs

# It's important to plot the average value for district so I create a new table with that information 
london_avg_p = london_df_2015 %>% group_by(district) %>% summarise(price = mean(price)) %>% arrange(price)

## House prices between Boroughs?
ggplot(data = london_df_2015, 
       aes(x = fct_relevel(district, rev(london_avg_p$district)), # In this line I indicate the order which is the reverse of the values in london_avg_p
           y = price)) + 
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3) + 
  geom_point(data = london_avg_p, aes(x = district, y = price, color = 'red'), size = 1) + 
  coord_flip() +
  scale_y_log10() + 
  labs(x = "London Boroughs", 
       y = "Log of Price", 
       color = "Mean Price") +
  ggtitle("House Prices by Borough: London 2015") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8))

```

The graph above shows that for 2015 the city of London was the borough with the highest price in housing. The City of Westminster as well of Kensington and Chelsea are other expensive boroughs of London. On the other hand, according to the data Barking and Dagenham is the borough where on average housing is less expensive.

A better most complete graph can show the 

```{r, warning=FALSE, message=F}

## Data frame for London with the columns year
london_df = london_df %>% mutate(Year_of_transfer = format(date_of_transfer, format = "%Y"))

## House prices between Boroughs?
ggplot(data = london_df %>% filter(Year_of_transfer %in% c(2012,2013,2014)), # Filtering london for the last previous years
       aes(x = fct_relevel(district, rev(london_avg_p$district)), # In this line I indicate the order which is the reverse of the values in london_avg_p
           y = price)) + 
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3) + 
  coord_flip() +
  scale_y_log10() + 
  facet_wrap(~Year_of_transfer, ncol=3) + 
  labs(x = "London Boroughs", 
       y = "Log of Price", 
       color = "Mean Price") +
  ggtitle("House Prices by Borough: London 2012 - 2014") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8))

```
From the years 2012 to 2014 using the median value of the sell price of house we can see that Kensington and Chelsea wes the most expensive Borough.

### Relationship between price and floor levels

We want to see the relationship between price and floor level, however the data do not have explicitly the information of floor level, for this reason I will use the SAON column which is a text column with the information of location in the case of properties that are divide into separate unites such as flats.

Note: Given that there is not a specific column with the information of floor level I will try to retrieve the information of the floor using regular expressions, this exercise is biased since depends on the way the information of floor is filled (there is not an specific format) and the methods of extraction created.

```{r, warning=FALSE, message=F}

### Create the floor level (only for flats (property_type == "F"))) by extracting the information from SAON

#### The next pipeline extracts as well as possible the information of the floor by the following proccess
  #### 1 Using the strings to denote floors from basement to 9th floor
  #### Taking the first digit in the case of floor numbers of two digits
  #### Taking the first two digits in the case of floor numbers of three digits
  #### Select up to 19 floors levels

floor_price_df = df %>% 
  filter(property_type == "F" & format(date_of_transfer, format = "%Y") == 2015) %>%
  mutate(Floor_x = str_replace_all(SAON, c(".*BASEMENT.*" = "-1",
                                         ".*GROUND.*" = "0",
                                         ".*FIRST.*" = "1",
                                         ".*SECOND*" = "2",
                                         ".*THIRD*" = "3",
                                         ".*FOURTH*" = "4",
                                         ".*FIFTH*" = "5",
                                         ".*SIXTH*" = "6",
                                         ".*SEVENTH*" = "7",
                                         ".*EIGHTH*" = "8",
                                         ".*NINETH*" = "9",
                                         "FLAT " = "",
                                         "APARTMENT " = ""))) %>%
  mutate(Floor_x = as.character(as.numeric(Floor_x))) %>% 
  mutate(Floor_x = case_when(nchar(Floor_x) == 1 ~ Floor_x,
                           nchar(Floor_x) == 2 & Floor_x != "-1" ~ str_sub(Floor_x, 1, 1),
                           nchar(Floor_x) == 3 ~ str_sub(Floor_x, 1, 2)
                           )) %>% 
  mutate(Floor = as.numeric(Floor_x)) %>%
  select(property_type, SAON, Floor, price) %>%
  filter(Floor < 20) %>%
  drop_na(Floor)

# Run a regression between floor level a price
lm1 = lm(log(price) ~ Floor, floor_price_df)
reg_text = paste("log(price) =", round(lm1$coefficients[1],2), "+", round(lm1$coefficients[2],3), "floor_level")

# Create a p graph with an annotation of the regression results
## In this kind of situations is better to work with jitter isntead of point (geom_point)

ggplot(data = floor_price_df, aes(x = Floor, y = price)) +
  scale_y_log10() +
  geom_jitter(alpha = 0.2, size = 0.5) +
  geom_smooth(method='lm', formula = y ~ x) + 
  annotate("text", x = 13, y = 1e+07, label = reg_text) +
  labs(x = "Floor level", 
       y = "Log of Price") +
  ggtitle("Floor level vs. price") +
  theme_minimal() + 

```

The graph above shows a very small association between floor level and the prices. The results of running a OLS regression indicate that on average as the floor level increases the housing price increases in 1.4%.



### Prices of the houses throughout the year (2015)

```{r, warning=FALSE, message=F}

df_2015 = df %>%
              mutate(year_transfer = format(date_of_transfer, format = "%Y"),
                     month_transfer = format(date_of_transfer, format = "%m")) %>% 
              filter(year_transfer == 2015)
              mutate(season = case_when(month_transfer %in% c("01", "02", "03") ~ "winter",
                                        month_transfer %in% c("04", "05", "06") ~ "spring",
                                        month_transfer %in% c("07", "08", "09") ~ "summer",
                                        month_transfer %in% c("10", "11", "12") ~ "fall"))

            
# Relation between month of the year and price
ggplot(data = london_df %>% filter(year_transfer == 2015), 
       aes(x = month_transfer, y = price)) +
  scale_y_log10() +
  geom_boxplot()


```




